Meta-analytic approaches and effect sizes to account for 'nuisance heterogeneity' in comparative physiology

### 2021-10-07 MS/ms.Rmd
small changes
Just some editing. Reword new section

Meta-analysis is a powerful tool to generate quantitatively informed answers to pressing global challenges. By distilling data from broad sets of research designs and study systems into standardized effect sizes, meta-analyses provide physiologists with opportunities to estimate overall effect sizes and understand the drivers of effect variability. Despite this ambition, research designs in the field of comparative physiology can appear, at the outset, as being vastly different to each other because of 'nuisance heterogeneity' (e.g., different temperatures or treatment dosages used across studies). Methodological differences across studies have led many to believe that meta-analysis is an exercise in comparing "apples with oranges". Here, we dispel this myth by showing how standardised effects sizes can be used in conjunction with multilevel meta-regression models to both account for the factors driving differences across studies and make them more comparable. We assess the prevelance of nuisance hetrogeneity in the comparative physiology literature -- showing it is common and often not accounted for in analyses. We then formalise effect size measures (e.g., the temperature coefficient, $Q_{10}$) that provide comparative physiologists with a means to remove nuisance heterogeneity without the need to resort to more complex statistical models that may be harder to interpret. We also describe more general approaches that can be applied to a variety of different contexts to derive new effects sizes and sampling variances, opening up new possibilities for quantitative synthesis. By using effect sizes that account for components of effect heterogeneity, in combination with existing meta-analytic models, comparative physiologists can explore exciting new questions and make results from large-scale data sets more accessible, comparable and widely interpretable.

@@

The goal of our review is to briefly overview the meta-analytic process and argue for the value of heterogeneity in answering fundamental research problems and guiding research directions. We present results from a survey of published meta-analyses in the field of comparative physiology to gauge methodological and meta-analytic practices -- determining what types of effect sizes are commonly used, how often effect size estimates are impacted by nuisance heterogeneity and whether meta-analytic models (i.e., models that account for sampling variance) are commonly applied. Then, we show how 'nuisance heterogeneity' in comparative physiology can be comfortably dealt with by re-formalising many existing effect size measures and / or by using multilevel meta-regression models. We formalise alternative effect sizes and their associated sampling variance to provide comparative physiologists with opportunities to explicitly incorporate 'nuisance heterogeneity' at the effect size level to ease their interpretation. Finally, we describe how more complex treatment differences, such as non-linear dosage differences, can be accommodated using multilevel meta-regression models. We hope that, by expanding the meta-analytic toolkit, it will provide new opportunities for comparative physiologists to address how organisms will cope with rapidly changing environments and anthropogenic stressors in the future.

@@

Meta-analysis is always concerned with effect heterogeneity; in other words, the factors that drive differences in the direction and magnitude of effects within and across studies [@Nakagawa2017; @Gurevitch2018; @Cooperetal2019; @Borenstein2019; @Lag2010]. The concept of heterogeneity is vitally important because it tells us how general our findings are likely to be and how much of the variance we see is the result of real biology or methodological differences (after meta-analytically accounting for sampling variance) [@Nakagawa2017; @Borenstein2019; @Lag2010].

@@

## **Nuisance Heterogeneity in Comparative Physiology**

Certain types of heterogeneity get in the way of understanding real biological patterns and are simply a ‘nuisance’ to interpreting effects. Take for example a meta-analysis that attempts to assess the effects of diet on fish growth across different populations of a widely distributed (cosmopolitan) species. Studies might rear populations of the same fish species under different temperatures, but in the end, they are still all governed by the same thermodynamic effects on growth. In this situation, we are still interested in understanding overall diet effects, but we may question whether such a conclusion is sensible given the diversity of temperatures applied. If temperature differences are of direct interest, then this variable can be formally incorporated in statistical models. If, however, it is not of primary interest, then it may introduce an unnecessary complication to the interpretation of the effect of diet on fish growth. Such heterogeneity is an important  contributing factor to the ‘apples and oranges’ problems described above.

>>>>

Meta-analysis is a powerful tool to generate quantitatively informed answers to pressing global challenges. By distilling data from broad sets of research designs and study systems into standardized effect sizes, meta-analyses provide physiologists with opportunities to estimate overall effect sizes and understand the drivers of effect variability. Despite this ambition, research designs in the field of comparative physiology can appear, at the outset, as being vastly different to each other because of 'nuisance heterogeneity' (e.g., different temperatures or treatment dosages used across studies). Methodological differences across studies have led many to believe that meta-analysis is an exercise in comparing "apples with oranges". Here, we dispel this myth by showing how standardised effects sizes can be used in conjunction with multilevel meta-regression models to both account for the factors driving differences across studies and make them more comparable. We assess the prevelance of nuisance hetrogeneity in the comparative physiology literature -- showing it is common and often not accounted for in analyses. We then formalise effect size measures (e.g., the temperature coefficient, $Q_{10}$) that provide comparative physiologists with a means to remove nuisance heterogeneity without the need to resort to more complex statistical models that may be harder to interpret. We also describe more general approaches that can be applied to a variety of different contexts to derive new effects sizes and sampling variances, opening up new possibilities for quantitative synthesis. By using effect sizes that account for components of effect heterogeneity, in combination with existing meta-analytic models, comparative physiologists can explore exciting new questions while making results from large-scale data sets more accessible, comparable and widely interpretable.

@@

The goal of our review is to briefly overview the meta-analytic process and argue for the value of heterogeneity in answering fundamental research problems and guiding research directions. We present results from a survey of published meta-analyses in the field of comparative physiology to gauge methodological and meta-analytic practices -- determining what types of effect sizes are commonly used, how often effect size estimates are impacted by nuisance heterogeneity and whether meta-analytic models (i.e., models that account for sampling variance) are commonly applied. Then, we show how 'nuisance heterogeneity' in comparative physiology can be comfortably dealt with by re-formalising many existing effect size measures and / or by using multilevel meta-regression models. We formalise alternative effect sizes and their associated sampling variance to provide comparative physiologists with opportunities to explicitly incorporate nuisance heterogeneity at the effect size level to ease their interpretation. Finally, we describe how more complex treatment differences, such as non-linear dosage differences, can be accommodated using multilevel meta-regression models. We hope that, by expanding the meta-analytic toolkit, it will provide new opportunities for comparative physiologists to address how organisms will cope with rapidly changing environments and anthropogenic stressors in the future.

@@

Meta-analysis is always concerned with effect heterogeneity; in other words, the factors that drive differences in the direction and magnitude of effects within and across studies [@Nakagawa2017; @Gurevitch2018; @Cooperetal2019; @Borenstein2019; @Lag2010]. The concept of heterogeneity is vitally important because it tells us how general our findings are likely to be and how much of the variance we see is the result of real biology or methodological differences (after accounting for sampling variance) [@Nakagawa2017; @Borenstein2019; @Lag2010].

@@

## **Nuisance Heterogeneity Complicates Meta-analytic Interpretations**

Nuisance heterogeneity can get in the way of understanding real biological patterns that interest comparative physiologists. Take for example a meta-analysis that attempts to assess the effects of diet on fish growth across different populations of a widely distributed (cosmopolitan) species. Studies might rear populations of the same fish species under different temperatures, but in the end, they are still all governed by the same thermodynamic effects on growth. In this situation, we are still interested in understanding overall diet effects, but we may question whether such a conclusion is sensible given the diversity of temperatures applied. If temperature differences are of direct interest, then this variable can be formally incorporated in statistical models. If, however, it is not of primary interest, then it may introduce an unnecessary complication to the interpretation of the effect of diet on fish growth. Such heterogeneity is an important  contributing factor to the ‘apples and oranges’ problems described above.
